---
title: "Fairness/causation: Stop and Frisk dataset"
output: github_document
bibliography: bibliography.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(forcats)
library(ggplot2)
library(reshape2)
library(e1071)
library(glmnet)
source("functions.R")
```

## Data

From the [NYC Stop and Frisk dataset](http://www.nyclu.org/content/stop-and-frisk-data) @stop:2014. 

```{r}
data <- read.csv("2014_stop_and_frisk.csv", strip.white = TRUE, header = TRUE)
# only use part of the data because of NAs, then remove NAs
data <- data[,which(apply(data, 2, function(col) mean(is.na(col))) < .1)]
data$year <- NULL
data$linecm <- NULL
data$dob <- NULL
data$datestop <- NULL
data$repcmd <- NULL
data$revcmd <- NULL
faccols <- which(apply(data, 2, function(col) length(unique(col))) < 10)
for (col in faccols) data[,col] <- factor(data[,col])
stringcols <- which(sapply(data[1,], function(v) is.factor(v) & nlevels(v) > 10))
data <- data[,-stringcols]
data <- data[complete.cases(data),]
data$pct <- factor(data$pct)
data$race <- fct_recode(factor(data$race),
                        Black = "1", BlaskHisp = "2", Hispanic = "3",
                        White = "4", AsianPacI = "5", NativeAm = "6")
data$arstmade <- as.numeric(data$arstmade)
meanarst <- mean(data$arstmade)
data$arstmade <- data$arstmade - meanarst
output <- data[,c("race", "sex", "arstmade")]
t(head(data))[,1:3]
```

## Imbalances in the data

Average outcomes by race.

```{r}
data %>% group_by(race) %>% summarise(count = n(), arstmade = mean(arstmade))
```

### Linear regression

Predicted averages coincide with data averages.

```{r}
model.lm <- lm(arstmade ~ .-1, data)
output$predlm <- predict(model.lm)
```

### Blinded to unfair covariates

Outcomes are only slightly less biased.

```{r}
lm.blind <- fairpred_blind(data, "arstmade", "race", method = "lm")
output$lmblind <- predict(lm.blind)
```

### Two-stage procedure

Greedily enforces fairness first, then builds predictions.

```{r}
lm.lm <- fairpred_2s(data, "arstmade", "race", method1 = "lm", method2 = "lm")
output$lm_lm <- predict(lm.lm)
```

### Comparing imbalance

Subgroup means.

```{r}
output[,c("race","predlm","lmblind","lm_lm")] %>% group_by(race) %>%
  summarise_if(.predicate = function(v) is.numeric(v), .funs = funs("mean"))
```

Distribution plots.

```{r}
pd <- melt(output)
ggplot(pd, aes(value, fill = sex)) + geom_density(alpha = .3) +
  facet_wrap(~variable) + xlim(-.5, 1) + theme_bw()

ggplot(pd, aes(value, fill = race)) + geom_density(alpha = .3) +
  facet_wrap(~variable) + xlim(-.5, 1) + theme_bw()
```

### Comparing (in-sample) mean-squared prediction error

```{r}
outputMSE <- data.frame((output$arstmade - output[,4:ncol(output)])^2)
outputMSE$const <- output$arstmade^2
outputMSE %>% summarise_all("mean")
```
